---
title: "Reef To Ridge Biodiversity Survey analyses"
format: pdf
editor: visual
---

## Introduction

We studied different aspects of the wildlife community along an altitudinal transect in southern Costa Rica. We started at the Osa Peninsula, close to sea level, and studied sites along the peninsula and into the mainland, going all the way to the mountains at the Parque Internacional La Amistad (PILA).

### Dung beetle survey

We collected dung beetles (Scarabeidae) using pitfall traps. Traps were baited with carrion or dung. We identified the dung beetles to the lowest taxonomic level (usually species or genus) and counted the number of individuals per species in every sample. We calculated richness as the total number of distinct species in a sample, and abundance as the total number of beetles of all species in the sample.

### Camera-trap survey

We placed motion-activated cameras along the linear transect. We set cameras 1 km apart from each other along the transect. Motion-activated cameras inform about the presence of different species of mammals, as well as relative abundance of the same species at different locations, or of different species (although the latter needs to be carefully interpreted).

```{r}

#|echo=FALSE
library(tidyverse)
library(corrplot)
```

#### Ordination

```{r}
sp_mat <- DAT %>% group_by(Site, Species) %>% count() %>% pivot_wider(names_from = Species, values_from = n, values_fill = 0) %>% ungroup() %>% select(-1) 
mds <- metaMDS(sp_mat)
plot(mds,  type = 'none', las=1)
ordiellipse(mds,env_mat[,3],draw = "polygon", col = hcl.colors(2))
# plot(envfit(mds, env_mat[,3]), col = 'darkred')
text(mds, display = 'species', cex = 0.8)
legend("bottomright", legend = c("PILA","Corcovado"),fill = hcl.colors(2))
```

#### JSDM

We fit a joint species distribution model to analyze the occupancy probability of different species, and how the expected occupancy is affected by environmental covariates such as land cover and connectivity.

```{r}
library(sjSDM)
DAT <- read_csv("../Data/combined_records.csv")
occ_data_bin <- DAT %>% 
  distinct(Site, Species) %>%
  mutate(n=1) %>% 
  pivot_wider(names_from = Species, values_from = n, values_fill = 0) %>% 
  column_to_rownames("Site") %>% 
  as.matrix()
env_mat <- tibble(Site = rownames(occ_data_bin)) %>%
  left_join(distinct(all_stn_effort, Site,lat,long)) %>% 
  mutate(area = ifelse(grepl(x = Site, pattern = "PILA"), 1,0)) %>% 
  column_to_rownames("Site") %>% 
  as.matrix()
occ_models <- list()
# fit a basic model with no biotic interactions or spatial autocorrelation.
occ_models[["basic"]] <- sjSDM(Y = occ_data_bin, env = linear(data = scale(env_mat), formula = ~area), # scales env. covariates: canopy, dist to road, dist to pa, forest cover (land use), evi
               se = TRUE, family=binomial("probit"), device = "cpu")
```

```{r}
plot(occ_models$basic)
```

```{r}
XY <- sf::sf_project("EPSG:4326","EPSG:5367",env_mat[,2:1])
colnames(XY) <- c("x","y")
occ_models[["spatial"]] <- sjSDM(Y = occ_data_bin, 
                                 env = linear(data = scale(env_mat), formula = ~area), # scales env. covariates: canopy, dist to road, dist to pa, forest cover (land use), evi
               spatial = linear(data = scale(XY), formula = ~0+x:y), 
               se = TRUE, 
               family=binomial("probit"), 
               device = "cpu")
plot(occ_models$spatial)
```

```{r}
spModAOV <- anova(occ_models$spatial)
plot(spModAOV, internal = T)
```

```{r}
occ_rsquared <- sapply(occ_models, Rsquared)
occ_loglik <- sapply(occ_models, logLik)
cbind(occ_loglik, occ_rsquared)
```

```{r}
bestmod <- occ_models$basic
bestmod_summary <- summary(bestmod)

bestmodcoefs <- data.frame(coef(bestmod)[[1]][,-1])
names(bestmodcoefs) <- bestmod$names[-1]
bestmodcoefs$sp <- bestmod$species
bestmodcoefs <- pivot_longer(bestmodcoefs, cols = 1, names_to = "coef", values_to = "value")
# significance
sigVals <- sigLabel <- data.frame(t(bestmod_summary$P)[,-1])
names(sigVals) <- names(sigLabel) <- bestmod$names[-1]
sigLabel[bestmod$weights[[1]][,-1]>0,1] <- "+"
sigLabel[bestmod$weights[[1]][,-1]<0,1] <- "-"
sigLabel[sigVals$area>0.05 | is.na(sigVals$area),] <- ""
# sigLabel <- replace(sigLabel,bestmod$weights[[1]][,-1]>0,values = "+") %>% 
#   replace(bestmod$weights[[1]][,-1]<0,values = "-") %>% 
#   replace(sigLabel>0.05, "")
sigLabel$sp <- bestmod$species
sigLabel <- pivot_longer(sigLabel, cols = 1, names_to = "coef", values_to = "text")

# plot
ggplot(bestmodcoefs, aes(x=coef, y=fct_reorder(sp, value)))+
  geom_tile(aes(fill = value))+
  scale_fill_gradient2(limits = c(-3,3))+
  # theme_void()+
  theme(axis.text.x = element_text(angle = 45, size = 9, hjust = 0), 
        # legend.position = 'none',
        axis.title = element_blank())+
  scale_x_discrete(expand = expansion(mult = c(0.2,0.5)), position = 'top', labels = c("Area"))+
  geom_text(data=left_join(bestmodcoefs,sigLabel), aes(label=text), vjust=0.5) -> jsdm_coef_plot
jsdm_coef_plot
```

```{r}
# arrange species by their dissimilarity for correlation matrix
sp_hclust_order <- hclust(dist(cov2cor(getCov(bestmod))))$order
# import common names database
common <- read.csv("../Data/species_commonName.csv")
common_hclust <- data.frame(common, order = sp_hclust_order)

corrmat <- matrix(nrow = length(sp_hclust_order), ncol = length(sp_hclust_order))
for (i in seq_along(sp_hclust_order)) {
  for (j in seq_along(sp_hclust_order)) {
    corrmat[i,j] <- cov2cor(getCov(bestmod))[sp_hclust_order[i],sp_hclust_order[j]]
  }
}

# base R plot
par(mar = c(2,15,10,5), las=2)
plot(seq_along(sp_hclust_order), seq_along(sp_hclust_order), 
     type = "n", 
     xaxt = "n", yaxt = "n",bty="n",
     xlab = "", ylab = "",asp=1)
image(1:33,1:33,corrmat[ncol(corrmat):1,], add = T, col = hcl.colors(100,"Blue-Red 3"), zlim=c(-1,1))
axis(2, at = 1:23, labels = common$common_name[sp_hclust_order])
axis(3, at = 23:1, labels = common$common_name[sp_hclust_order])

## ggplot
corr_ggdf <- expand.grid(common$common_name[sp_hclust_order], common$common_name[sp_hclust_order])
for (i in seq_along(corrmat)) {
  corr_ggdf[i,3] <- corrmat[i]
}

names(corr_ggdf) <- c("sp1", "sp2", "corr")
col2 <- colorRampPalette(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061"))
ggplot(corr_ggdf, aes(sp1,sp2))+
  geom_tile(aes(fill = corr))+
    scale_fill_gradientn(limits=c(-1,1),
                         colours = col2(200))+
    labs(x='',y='', fill='')+
    theme(axis.text.y=element_text(hjust = 0.5), 
          axis.text.x=element_text(angle = 45, hjust = 0),
          legend.key.height = unit(3, 'lines'))+
    scale_x_discrete(position = 'top', limits = rev) -> jsdm_corr_plot
jsdm_corr_plot
```

### Soil chemistry

We collected soil samples at different locations along the transect. The samples were analyzed for their chemical composition. This included quantification of organic matter (carbon, nitrogen), water content and pH, as well as inorganic elements (Iron, Calcium, Copper, Magnesium, Manganese, Phosporous, Potassium, Zinc). We collected and analyzed 133 soil samples.

```{r}
soilchem <- read_csv("../Data/soil_chemistry.csv")
soilchem %>% pivot_longer(cols = where(is.numeric), names_to = "variable", values_to = "value") %>% group_by(variable) %>% summarise(mean = mean(value, na.rm = T), min = min(value), max = max(value), sd = sd(value, na.rm=T))
```

```{r, fig.cap="Correlation of soil chemistry variables"}
corrplot(cor(soilchem[,-1], use = "complete.obs"), order = "FPC")
```

There are strong positive correlations between the content of Mg, Ca, the soil pH and cationic exchange capacity. There also strong positive correlations between the content of C, N, P, and SA. These two groups are negatively correlated to each other. Elements like Zn, K, Cu show no strong association with other variables.

### Beetles

We also collected beetles (Coleoptera) using pit fall traps near each camera-trap location. Each location was sampled 1-3 times.

```{r}
beet_data <- read.csv("../Data/dung_beetles.csv")
names(beet_data)
beet_sp_mat <- beet_data %>% group_by(Plot) %>% summarise(across(6:72, sum)) %>%  column_to_rownames(var = "Plot") 
beet_sp_mat <- beet_sp_mat[rowSums(beet_sp_mat)>0,]
beet_sp_mat <- beet_sp_mat[rownames(beet_sp_mat)!="Ms134",]
mds_beet <- metaMDS(beet_sp_mat)
plot(mds_beet,  type = 'none', las=1)
ordiellipse(mds_beet,as.factor(grepl("P",rownames(beet_sp_mat))),draw = "polygon", col = hcl.colors(2))
# plot(envfit(mds, env_mat[,3]), col = 'darkred')
text(mds_beet, display = 'species', cex = 0.8)
# legend("bottomright", legend = c("PILA","Corcovado"),fill = hcl.colors(2))
```
