---
title: "Reef To Ridge Biodiversity Survey analyses"
format: pdf
editor: visual
---

## Introduction

We studied different aspects of the wildlife community along an altitudinal transect in southern Costa Rica. We started at the Osa Peninsula, close to sea level, and studied sites along the peninsula and into the mainland, going all the way to the mountains at the Parque Internacional La Amistad (PILA).

### Dung beetle survey

We collected dung beetles (Scarabeidae) using pitfall traps. Traps were baited with carrion or dung. We identified the dung beetles to the lowest taxonomic level (usually species or genus) and counted the number of individuals per species in every sample. We calculated richness as the total number of distinct species in a sample, and abundance as the total number of beetles of all species in the sample.

### Camera-trap survey

We placed motion-activated cameras along the linear transect. We set cameras 1 km apart from each other along the transect. Motion-activated cameras inform about the presence of different species of mammals, as well as relative abundance of the same species at different locations, or of different species (although the latter needs to be carefully interpreted).

```{r}

#|echo=FALSE
library(tidyverse)
library(corrplot)
```

### Soil chemistry

We collected soil samples at different locations along the transect. The samples were analyzed for their chemical composition. This included quantification of organic matter (carbon, nitrogen), water content and pH, as well as inorganic elements (Iron, Calcium, Copper, Magnesium, Manganese, Phosporous, Potassium, Zinc). We collected and analyzed 133 soil samples.

```{r}
soilchem <- read_csv("../Data/soil_chemistry.csv")
soilchem %>% pivot_longer(cols = where(is.numeric), names_to = "variable", values_to = "value") %>% group_by(variable) %>% summarise(mean = mean(value, na.rm = T), min = min(value), max = max(value), sd = sd(value, na.rm=T))
```

#### Variable Correlation

```{r, fig.cap="Correlation of soil chemistry variables"}
corrplot(cor(soilchem[,-1], use = "complete.obs"), order = "FPC")
```

There are strong positive correlations between the content of Mg, Ca, the soil pH and cationic exchange capacity. There also strong positive correlations between the content of C, N, P, and SA. These two groups are negatively correlated to each other. Elements like Zn, K, Cu show no strong association with other variables.

#### PCA

Since there are many variables it makes sense to summarize them using principal components. We scale since the variables vary orders of magnitude. We see, however, that the data is very heterogeneous, and cannot be summarized neatly in a few components.

```{r, echo=FALSE}
par(mfrow = c(1,2))
biplot(princomp(scale(soilchem[complete.cases(soilchem),-1])))
screeplot(princomp(scale(soilchem[complete.cases(soilchem),-1])), main="")
```
